---
title: "CMDA-4654  \n Project 1"
author: "Kathleen Reece, Curtis Wong, and Abia Zahir"
date: "`r format(Sys.time(),'%B %d, %Y')`"
output:
  pdf_document:
    highlight: haddock
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---
  
<!-- The above is set to automatically compile to a .pdf file.   -->
<!-- It will only succeed if LaTeX is installed. -->


```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, # Required
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE # Turn off warnings
                      )

```

```{r}
library(ggplot2)
library(tidyverse)
library(kableExtra)
library(patchwork)
source("Reece_Wong_Zahir_Project_1.R")
```


\clearpage
<!-- ---------------------------------------------------------------------------------------------------- -->
<!-- ---------------- Homework Problems start below these lines ----------------------------------------- -->
<!-- ---------------------------------------------------------------------------------------------------- -->


# Problem 1

```{r}
# Loading in the data set
load("ozone.RData")
data("ozone")
```

## Part 1

```{r}
# Fit models and overlay on top of each other
# For easy comparison
ggplot(ozone, aes(x = temperature, y = ozone)) + theme_bw() + geom_point() +
  geom_smooth(method = "lm", se = F, aes(color = "1st")) +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 2)", se = F, aes(color = "2nd")) +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 3)", se = F, aes(color = "3rd")) +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 4)", se = F, aes(color = "4th")) +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 5)", se = F, aes(color = "5th")) +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 6)", se = F, aes(color = "6th")) +
  scale_color_manual(name = "Polynomial Degree", 
                     breaks = c("1st", "2nd", "3rd", "4th", "5th", "6th"), 
                     values = c("red", "orange", "yellow", "green", "blue", "purple")) +
  labs(title = "Polynomial Regression of Ozone on Temperature")
```

From the graph, it seems like the 2nd-degree fit is the best. However, none of the degrees seem to give a satisfactory fit.

## Part 2

```{r}
# Where we'll store all the LOESS fits we're making
models <- vector(mode = "list", length = 22)
# Staring span value
span_val <- 0.25
# Fit all the degree-1 models
for (i in 1:11) {
  models[[i]] <- myloess(x_train = ozone$temperature, y_train = ozone$ozone,
                       span = span_val, show.plot = FALSE)
  span_val <- span_val + 0.05
}
# Reset span
span_val <- 0.25
# Fit all the degree-2 models
for (i in 12:22) {
  models[[i]] <- myloess(x_train = ozone$temperature, y_train = ozone$ozone,
                       span = span_val, degree = 2, show.plot = FALSE)
  span_val <- span_val + 0.05
}
# Put the span, degree, and MSE into dataframe for easy table making
model_df <- dplyr::bind_rows( lapply(models, function(m) {
  data.frame(
      span = m$span,
      degree = m$degree,
      MSE = m$MSE
    )
  })
)
# Output table
knitr::kable(
  model_df,
  caption = "LOESS Fits on Ozone Data"
)
```

From the table, the best fits with a degree of 1 occur for span values of 0.25, 0.30, and 0.35, and the best fits with a degree of 2 occur for span values of 0.25, 0.30, and 0.35 as well.

```{r}
# Use the plots from the myloess function, but add titles
p1 <- models[[1]]$loessplot + labs(title = "degree = 1, span = 0.25") 
p2 <- models[[2]]$loessplot + labs(title = "degree = 1, span = 0.30")
p3 <- models[[3]]$loessplot + labs(title = "degree = 1, span = 0.35")
p4 <- models[[11]]$loessplot + labs(title = "degree = 2, span = 0.25")
p5 <- models[[12]]$loessplot + labs(title = "degree = 2, span = 0.30")
p6 <- models[[13]]$loessplot + labs(title = "degree = 2, span = 0.35")
# Use patchwork library to plot all together
(p1 + p2 + p3) / (p4 + p5 + p6) + plot_annotation(title = "Best LOESS fits on Ozone data")
```

Based the plots, I would consider all 6 of the models to be better fits than any of the polynomial fits. However, I would also consider almost all of the curves to be somewhat overfitted to the data. In particular, the curves where the local fits used a degree 2 polynomial are very jagged, seemingly in response to some small shifts in the data; the degree-1 models with spans of 0.25 and 0.30 also show some of this jaggedness, though on a lesser degree. Of all of the "best" fits, the one with the least signs of overfitting is the LOESS curve with degree = 1 and span = 0.35.

## Part 3

```{r}
# Where we'll store all the LOESS fits we're making
baser_models <- vector(mode = "list", length = 22)
# Staring span value
span_val <- 0.25
# Fit all the degree-1 models
for (i in 1:11) {
  baser_models[[i]] <- loess(ozone ~ temperature, data = ozone, span = span_val, degree = 1)
  span_val <- span_val + 0.05
}
# Reset span
span_val <- 0.25
# Fit all the degree-2 models
for (i in 12:22) {
  baser_models[[i]] <- loess(ozone ~ temperature, data = ozone, span = span_val)
  span_val <- span_val + 0.05
}
# Put the span, degree, and RSE into dataframe for easy table making
baser_model_df <- dplyr::bind_rows(lapply(baser_models, function(m) {
  data.frame(
      span = m$pars$span,
      degree = m$pars$degree,
      RSE = m$s
    )
  })
)
# Output table
knitr::kable(
  baser_model_df,
  caption = "Loess Fits on Ozone Data, with loess()"
)
```

Although the built in `loess()` function uses RSE instead of MSE, it agrees on the best 3 models of each degree.

```{r}
# Use the plots from the myloess function, but add titles
g1 <- ggplot(ozone, aes(x = temperature, y = ozone)) + geom_point() + theme_bw() +
  geom_smooth(method = "loess", span = 0.25, se = F, method.args = list(degree = 1)) +
  labs(title = "degree = 1, span = 0.25")
g2 <- ggplot(ozone, aes(x = temperature, y = ozone)) + geom_point() + theme_bw() +
  geom_smooth(method = "loess", span = 0.3, se = F, method.args = list(degree = 1)) +
  labs(title = "degree = 1, span = 0.30")
g3 <- ggplot(ozone, aes(x = temperature, y = ozone)) + geom_point() + theme_bw() +
  geom_smooth(method = "loess", span = 0.35, se = F, method.args = list(degree = 1)) +
  labs(title = "degree = 1, span = 0.35")
g4 <- ggplot(ozone, aes(x = temperature, y = ozone)) + geom_point() + theme_bw() +
  geom_smooth(method = "loess", span = 0.25, se = F, method.args = list(degree = 2)) +
  labs(title = "degree = 2, span = 0.25")
g5 <- ggplot(ozone, aes(x = temperature, y = ozone)) + geom_point() + theme_bw() +
  geom_smooth(method = "loess", span = 0.3, se = F, method.args = list(degree = 2)) +
  labs(title = "degree = 2, span = 0.30")
g6 <- ggplot(ozone, aes(x = temperature, y = ozone)) + geom_point() + theme_bw() +
  geom_smooth(method = "loess", span = 0.35, se = F, method.args = list(degree = 2)) +
  labs(title = "degree = 2, span = 0.35")
# Use patchwork library to plot all together
(g1 + g2 + g3) / (g4 + g5 + g6) + plot_annotation(title = "Best LOESS fits on Ozone data, with loess()")
```

The plots look very similar to those from the `myloess()` function. The only difference is that the curves look a tad smoother.

# Problem 2

```{r}
# Loading the data
library(MASS)
data("mcycle")
load("mcycle_test.RData")
data("mcycle_test")
```

## Part 1

We can use the same fitting-and-plotting process as in part 3 of problem 1:

```{r}
# Where we'll store all the LOESS fits we're making
models <- vector(mode = "list", length = 22)
# Staring span value
span_val <- 0.25
# Fit all the degree-1 models
for (i in 1:11) {
  models[[i]] <- myloess(x_train = mcycle$times, y_train = mcycle$accel,
                       span = span_val, show.plot = FALSE)
  span_val <- span_val + 0.05
}
# Reset span
span_val <- 0.25
# Fit all the degree-2 models
for (i in 12:22) {
  models[[i]] <- myloess(x_train = mcycle$times, y_train = mcycle$accel,
                       span = span_val, degree = 2, show.plot = FALSE)
  span_val <- span_val + 0.05
}
# Put the span, degree, and MSE into dataframe for easy table making
model_df <- dplyr::bind_rows( lapply(models, function(m) {
  data.frame(
      span = m$span,
      degree = m$degree,
      MSE = m$MSE
    )
  })
)
# Output table
knitr::kable(
  model_df,
  caption = "LOESS Fits on mcycle Data"
)
```

For degree = 1, the best span values are 0.25, 0.30, and 0.35. For degree = 2, the best span values are also 0.25, 0.30, and 0.35.

```{r}
# Use the plots from the myloess function, but add titles
p1 <- models[[1]]$loessplot + labs(title = "degree = 1, span = 0.25") 
p2 <- models[[2]]$loessplot + labs(title = "degree = 1, span = 0.30")
p3 <- models[[3]]$loessplot + labs(title = "degree = 1, span = 0.35")
p4 <- models[[11]]$loessplot + labs(title = "degree = 2, span = 0.25")
p5 <- models[[12]]$loessplot + labs(title = "degree = 2, span = 0.30")
p6 <- models[[13]]$loessplot + labs(title = "degree = 2, span = 0.35")
# Use patchwork library to plot all together
(p1 + p2 + p3) / (p4 + p5 + p6) + plot_annotation(title = "Best LOESS fits on mcycle data")
```

All of the plots look pretty good, but in my opinion, the models with degree = 2 do a better job capturing the true shape than the models with degree = 1, especially for the flat section of points at the far left and the large curve in the middle of the plot. Also, all of the best models with degree = 2 have lower MSE values than all of the best models with degree = 1. Thus, I would say LOESS with degree = 2 and a span of 0.25, 0.30, or 0.35 provides the best fits to the data.

## Part 2

```{r}
# Where we'll store all the LOESS fits we're making
test_models <- vector(mode = "list", length = 24)
# Staring span value
span_val <- 0.1
# Fit all the degree-1 models from span = 0.1 to 0.65, incrementing by 0.05
for (i in 1:12) {
  test_models[[i]] <- myloess(x_train = mcycle$times, y_train = mcycle$accel,
                              x_test = mcycle_test$times, y_test = mcycle_test$accel,
                       span = span_val, show.plot = FALSE)
  span_val <- span_val + 0.05
}
# Reset span
span_val <- 0.1
# Fit all the degree-2 models from span = 0.1 to 0.65, incrementing by 0.05
for (i in 13:24) {
  test_models[[i]] <- myloess(x_train = mcycle$times, y_train = mcycle$accel,
                              x_test = mcycle_test$times, y_test = mcycle_test$accel,
                       span = span_val, degree = 2, show.plot = FALSE)
  span_val <- span_val + 0.05
}
# Put the span, degree, and MSE into a dataframe for easy table making
test_model_df <- dplyr::bind_rows( lapply(test_models, function(m) {
  data.frame(
      span = m$span,
      degree = m$degree,
      MSE = m$MSE
    )
  })
)
# Output table
knitr::kable(
  test_model_df,
  caption = "LOESS Fits on mcycle testing data"
)
```

For degree = 1, the optimal span that minimizes MSE seems to be around 0.10., and for degree = 2, the optimal span that minimizes MSE seems to be around 0.20. We can test spans closer to these values:

```{r}
# Where we'll store all the LOESS fits we're making
test_models2 <- vector(mode = "list", length = 18)
# Staring span value
span_val <- 0.06
# Fit all the degree-1 models from span = 0.06 to 0.14, incrementing by 0.01
for (i in 1:9) {
  test_models2[[i]] <- myloess(x_train = mcycle$times, y_train = mcycle$accel,
                              x_test = mcycle_test$times, y_test = mcycle_test$accel,
                       span = span_val, show.plot = FALSE)
  span_val <- span_val + 0.01
}
# Reset span
span_val <- 0.16
# Fit all the degree-2 models from span = 0.16 to 0.24, incrementing by 0.01
for (i in 10:18) {
  test_models2[[i]] <- myloess(x_train = mcycle$times, y_train = mcycle$accel,
                              x_test = mcycle_test$times, y_test = mcycle_test$accel,
                       span = span_val, degree = 2, show.plot = FALSE)
  span_val <- span_val + 0.01
}
# Put the span, degree, and MSE into a dataframe for easy table making
test_model_df <- dplyr::bind_rows( lapply(test_models2, function(m) {
  data.frame(
      span = m$span,
      degree = m$degree,
      MSE = m$MSE
    )
  })
)
# Output table
knitr::kable(
  test_model_df,
  caption = "LOESS Fits on mcycle testing data"
)
```

From our closer look, the optimal span when degree = 1 is 0.10, and the optimal span when degree = 2 is 0.17.

We can plot the data (testing and training) with the curves with the optimal spans for each degree

```{r}
best_span1 <- test_models2[[5]]$loessplot + 
  geom_point(data = mcycle, aes(x = times, y = accel), color = "gray") + 
  geom_point(color = "red") + labs(title = "degree = 1, span = 0.10")
best_span2 <- test_models2[[11]]$loessplot + 
  geom_point(data = mcycle, aes(x = times, y = accel), color = "gray") +
  geom_point(color = "red") + labs(title = "degree = 2, span = 0.17")

(best_span1 + best_span2) + plot_annotation("LOESS fits with spans minimizing MSE on testing mcycle data")
```

We can compare our plots to the results from R's built in `loess()` function:

```{r}
compare_d1 <- loess(accel ~ times, data = mcycle, span = 0.1, degree = 1)
compare_d2 <- loess(accel ~ times, data = mcycle, span = 0.17)
ypred_d1 <- predict(compare_d1, newdata = mcycle_test)
ypred_d2 <- predict(compare_d2, newdata = mcycle_test)
best_span1_baser <- ggplot(data = mcycle, aes(x = times, y = accel)) + geom_point(color = "gray") +
  geom_point(data = mcycle_test, aes(x = times, y = accel), color = "red") + theme_bw() +
  geom_line(data = data.frame(mcycle_test$times, ypred_d1), aes(x = mcycle_test$times, y = ypred_d1)) 
best_span2_baser <- ggplot(data = mcycle, aes(x = times, y = accel)) + geom_point(color = "gray") +
  geom_point(data = mcycle_test, aes(x = times, y = accel), color = "red") + theme_bw() +
  geom_line(data = data.frame(mcycle_test$times, ypred_d2), aes(x = mcycle_test$times, y = ypred_d2))
(best_span1_baser + best_span2_baser) + plot_annotation("LOESS fits (base R) with spans minimizing MSE on testing mcycle data")
```

# Problem 3

# Problem 4

## Part 1

## kNN

### Part 2

### Part 3

### Part 4

### Part 5

## dnkNN

### Part 2

### Part 3

### Part 4

### Part 5

# Problem 5

## Part a

## Part b

# Problem 6



